A quantitative Mac Diarmid's inequality for geometrically ergodic Markov chains

### 2019-12-13 Note_exponential_moment/MarkovCon.tex
update

\section{Applications to Unadjusted Langevin Algorithm}\label{sec:ULA}
We illustrate the applicability of the results by the example on one of the MCMC methods. Suppose that we aim at sampling from the distribution with density $\pi(x) = \frac{e^{-U(x)}}{\int\limits_{\mathbb{R}^d} e^{-U(y)}\,dy}$. One of the popular algorithms for solving this problem is the Unadjusted Langevin Algorithm, which suggests constructing the Markov Chain
$$
X_{k+1} = h_{k+1}(X_k) + \sqrt{2\gamma_{k+1}}Z_{k+1}, h_{k+1}(X_k) := X_{k} - \gamma_{k+1}\nabla U(X_{k}) 
$$
with some non-increasing sequence $\gamma_k \geq 0$, where $Z_k \sim \mathcal{N}(0,I)$ are i.i.d. standard normal $d$-dimensional random variables. Standard assumptions on the sequence $\{\gamma_k\}$ are $\sum\limits_{k=0}^{\infty}\gamma_k = \infty$ and $\sum\limits_{k=0}^{\infty}\gamma_k^2 < \infty$. Thus it is impossible to analyze properties of additive functionals of $\{X_k\}$ by standard tools, since the assumptions {\bf H2} and {\bf H3} fails in case $\gamma_k \rightarrow 0$.  Yet we might establish some alternative bounds on the speed of convergence to stationary distribution. We need to impose condition
\par
[{\bf C1}] Potential $U(x)$ satisfies for all $x,y \in \mathbb{R}^d$ the condition
$$
\|U(x) - U(y)\| \leq L\|x-y\|
$$
This condition obviously implies
$$
\|h_{k}(X) - h_k(Y)\| \leq (1+L\gamma_k)\|X-Y\|
$$
Hence, by \cite[Theorem~19]{durmus:moulines:2018}, we might obtain the following minorization condition
\begin{lemma} Assume that the condition {\bf C1} holds. Then for all $x,y \in \mathbb{R}^d, n\geq 1$ it holds
$$
\|\delta_xQ^n - \delta_yQ^n\|_{TV} \leq \mathbb{I}(x \neq y)\left(1 - 2\Phi\left(-\frac{\|x-y\|}{2\sqrt{b_n}}\right)\right)
$$
where $b_n := \sum\nolimits_{i=1}^{k}\frac{\gamma_i}{\prod_{j=1}^{i}(1+L\gamma_j)^2}$
\end{lemma}
Now we aim at establishing Foster-Lyapunov drift conditions for the kernel $P_{\gamma_n}$.We will say that function $V(x): \mathbb{R}^d \rightarrow [1,+\infty)$ satisfies a Foster-Lyapunov drift condition for the Markov kernel $P_\gamma$ if. there are such $\overline{\gamma} > 0, \lambda \in [0,1), C > 0$ such that for any $\gamma \in (0, \overline{\gamma}]$ it holds
\begin{equation}
\label{eq:drift}
P_{\gamma}V \leq \lambda^{\gamma}V + \gamma C
\end{equation}
Given that this condition holds, we may prove drift condition for. the kernel $Q^n$.
\begin{lemma} Assume that $P_{\gamma}$ satisfies condition \ref{eq:drift} for any $0 < \gamma < \overline{gamma}$. Then it holds
$$
Q^{n,p}V \leq \lambda^{\Gamma_{p,n}}V + C\frac{\lambda^{\Gamma_{1,p}} - \lambda^{\Gamma_{1,n}}}{\log{\frac{1}{\lambda}}}
$$ 
\end{lemma}
\begin{proof} For notation simplicity let us consider only $p=1$. Then by induction
$$
Q^{n}V \leq \lambda^{\Gamma_{1,n}}V + C \sum\limits_{k=1}^{n}\gamma_k \lambda^{\Gamma_{1,k-1}} \leq \lambda^{\Gamma_{1,n}}V + C \int\limits_{0}^{\Gamma_{1,n}}\lambda^{x}\,dx \leq
$$
$$
\leq \lambda^{\Gamma_{1,n}}V + C\frac{1-\lambda^{\Gamma_{1,n}}}{\log{\frac{1}{\lambda}}}
$$
\end{proof}

>>>>

\section{Applications to Unadjusted Langevin Algorithm}\label{sec:ULA}
We illustrate the applicability of the results by the example on one of the MCMC methods. Suppose that we aim at sampling from the distribution with density $\pi(x) = \frac{e^{-U(x)}}{\int\limits_{\mathbb{R}^d} e^{-U(y)}\,dy}$. One of the popular algorithms for solving this problem is the Unadjusted Langevin Algorithm, which suggests constructing the Markov Chain
$$
X_{k+1} = h_{k+1}(X_k) + \sqrt{2\gamma_{k+1}}Z_{k+1}, h_{k+1}(X_k) := X_{k} - \gamma_{k+1}\nabla U(X_{k}) 
$$
with some non-increasing sequence $\gamma_k \geq 0$, where $Z_k \sim \mathcal{N}(0,I)$ are i.i.d. standard normal $d$-dimensional random variables. Let us denote ULA kernel with parameter $\gamma$ as $P_{\gamma}$, that is,
\begin{equation}
\label{eq:ula_kernel}
P_{\gamma}(x,dy) = \frac{1}{(4\pi\gamma)^{\frac{d}{2}}}\exp{\left(-\frac{\|y-x+\gamma\nabla U(x)\|^2}{4\gamma}\right)}\,dy
\end{equation}
Let us also introduce the multiple steps kernel
$$
Q^{p,n} = P_{\gamma_p}\ldots P_{\gamma_n}, Q^{n} := Q^{1,n}
$$
Standard assumptions on the sequence $\{\gamma_k\}$ are $\sum\limits_{k=0}^{\infty}\gamma_k = \infty$ and $\sum\limits_{k=0}^{\infty}\gamma_k^2 < \infty$. Thus it is impossible to analyze properties of additive functionals of $\{X_k\}$ by standard tools, since the assumptions {\bf H2} and {\bf H3} fails in case $\gamma_k \rightarrow 0$.  Yet we might establish some alternative bounds on the speed of convergence to stationary distribution. We need to impose condition
\par
[{\bf C1}] Potential $U(x)$ satisfies for all $x,y \in \mathbb{R}^d$ the condition
$$
\|U(x) - U(y)\| \leq L\|x-y\|
$$
This condition obviously implies
$$
\|h_{k}(X) - h_k(Y)\| \leq (1+L\gamma_k)\|X-Y\|
$$
Hence, by \cite[Theorem~19]{durmus:moulines:2018}, we might obtain the following minorization condition
\begin{lemma} Assume that the condition {\bf C1} holds. Then for all $x,y \in \mathbb{R}^d, n\geq 1$ it holds
$$
\|\delta_xQ^n - \delta_yQ^n\|_{TV} \leq \mathbb{I}(x \neq y)\left(1 - 2\Phi\left(-\frac{\|x-y\|}{2\sqrt{b_n}}\right)\right)
$$
where $b_n := \sum\nolimits_{i=1}^{k}\frac{\gamma_i}{\prod_{j=1}^{i}(1+L\gamma_j)^2}$
\end{lemma}
Now we aim at establishing Foster-Lyapunov drift conditions for the kernel $P_{\gamma_n}$.We will say that function $V(x): \mathbb{R}^d \rightarrow [1,+\infty)$ satisfies a Foster-Lyapunov drift condition for the Markov kernel $P_\gamma$ if. there are such $\overline{\gamma} > 0, \lambda \in [0,1), C > 0$ such that for any $\gamma \in (0, \overline{\gamma}]$ it holds
\begin{equation}
\label{eq:drift}
P_{\gamma}V \leq \lambda^{\gamma}V + \gamma C
\end{equation}
Given that this condition holds, we may prove drift condition for. the kernel $Q^n$.

\begin{lemma}
\label{lem:drift_prod}
 Assume that $P_{\gamma}$ satisfies condition \ref{eq:drift} for any $0 < \gamma < \overline{gamma}$. Then it holds
$$
Q^{n,p}V \leq \lambda^{\Gamma_{p,n}}V + C\frac{\lambda^{\Gamma_{1,p}} - \lambda^{\Gamma_{1,n}}}{\log{\frac{1}{\lambda}}}
$$ 
\end{lemma}
\begin{proof} For notation simplicity let us consider only $p=1$. Then by induction
$$
Q^{n}V \leq \lambda^{\Gamma_{1,n}}V + C \sum\limits_{k=1}^{n}\gamma_k \lambda^{\Gamma_{1,k-1}} \leq \lambda^{\Gamma_{1,n}}V + C \int\limits_{0}^{\Gamma_{1,n}}\lambda^{x}\,dx \leq
$$
$$
\leq \lambda^{\Gamma_{1,n}}V + C\frac{1-\lambda^{\Gamma_{1,n}}}{\log{\frac{1}{\lambda}}}
$$
\end{proof}
Let us check that ULA kernel \ref{} satisfies the drift condition \label{eq:drift}. We will need additional assumption
\par
[{\bf C2}] There exists such constants $K_1 > 0$ and $m > 0$, such that for any $x \notin B(0,K_1)$ it holds
$$
\langle \nabla U(x), x \rangle \geq \frac{m}{2}\|x\|^2
$$
Another popular condition on the potential is strong convexity outside some euclidean ball:
\par
[{\bf C3}] There exist constants $K_1 > 0$ and $m > 0$ such that for any $x \notin B(0,K_1)$ and $y \in \mathbb{R}^d$, it holds $\langle D^2U(x)y,y\rangle \geq m\|y\|^2$. Moreover, there exists $M \geq 0$ such that for any $x \in \mathbb{R}^d$, $\|D^3U(x)\| \leq M$.
\par
Yet it is possible to show that assumption {\bf C3} and {\bf C1} imply {\bf C2} (see \cite[Lemma~14]{brosse:2018:diffusion}), so in the sequel we are going to work with the condition {\bf C2}. It allows us to prove the following result

\begin{lemma} Assume that the potential $U(x)$ satisfies conditions {\bf C1} and {\bf C2}. Then the kernel \ref{} satisfies drift condition \ref{eq:drift} for any $0 < \gamma < \overline{\gamma} = \frac{m}{4L^2}$ with drift function $V(x) = 1 + \|x\|^2$, constants $\lambda = \exp{-\left(\frac{m}{2}\right)}$, $C = 3K_1^2 + 2d + C_1$ for some absolute constant $C_1$.
\end{lemma}

\begin{proof} Consider $V(x) = 1 + \|x\|^2$, then
$$
P_{\gamma}V(x) = \int\limits_{\mathbb{R}^d}V(y)P_{\gamma}(x,dy) = \int\limits_{\mathbb{R}^d}(1+\|y\|^2)\frac{1}{(4\pi\gamma)^{\frac{d}{2}}}\exp{\left(-\frac{\|y-x+\gamma\nabla U(x)\|^2}{4\gamma}\right)}\,dy = 
$$

$$
=1 + \frac{1}{(4\pi\gamma)^{\frac{d}{2}}}\int\limits_{\mathbb{R}^d}\|z + x - \gamma \nabla U(x)\|^2 \exp{\left(-\frac{\|z\|^2}{4\gamma}\right)}\,dz
$$
Let us first consider the case $x \notin B(0,K_1)$. Note that $\|z+x-\gamma \nabla U(x)\|^2 = \|z\|^2 + 2 \langle z, x - \gamma \nabla U(x) \rangle + \gamma^2 \|x - \gamma \nabla U(x)\|^2$ and the linear term vanishes after integration, moreover, for $Z \sim \mathcal{N}(0,2\gamma I_d$, it holds $\mathbb{E}\|Z\|^2 = 2\gamma d$. It remains to notice that due to {\bf C1} and {\bf C2}
$$
\|x - \gamma \nabla U(x) \|^2 = \|x\|^2 - 2\gamma \langle \nabla U(x), x\rangle + \gamma^2 \|\nabla U(x)\|^2 \leq
$$
$$
 \leq (1-\gamma m)\|x\|^2 + 2\gamma^2 L^2 \|x\|^2 + 2\|\nabla U(0)\|^2
$$
Thus, plugging everything into expression for $P_{\gamma}V$ and using $\gamma < \frac{m}{4L}$, we obtain
$$
P_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + 2\gamma d + (\gamma m - 2\gamma^2 L^2) + 2\gamma^2\|\nabla U(0)\|^2 \leq \exp^{-\frac{\gamma m}{2}}V(x) + \gamma(d + 2\gamma \|\nabla U(0)\|^2)
$$
Now let $x \in B(0,K_1)$. Then simply using $\|x-\gamma \nabla U(x)\|^2 \leq 2(1+L\gamma)^2\|x\|^2 + 2\gamma^2\|\nabla U(0)\|^2$, we obtain
$$
P_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + \gamma\left((m - 2\gamma L^2)(1+\|x\|^2) + 2d + 2(1+L\gamma)^2\|x\|^d + 2\gamma \|\nabla U(0)\|^2\right) \leq 
$$

$$
\leq \exp^{-\frac{\gamma m}{2}}V(x) + \gamma(3K_1^2 + 2d + C_1)
$$
for some absolute constant $C_1$, which does not depend on $\gamma, d, x$.
\end{proof}

It is possible to use \ref{lem:drift_prod} in order to obtain convergence rates in $V$-norm, following the ideas of \cite[Theorem~18.4.3]{douc:moulines:priouret:soulier:2018}. Namely, the following lemma holds:
